A primer to numerical simulations: The perihelion motion of Mercury

### 2018-01-09 tex/MercuryNumerical.tex
C.H. modifications. Smoothening out beta part.

\usepackage[dvipsnames]{xcolor}         % Enabling colors by their 'svgnames'
\usepackage[%
  unicode=false,%
  pdftitle={A primer to numerical simulations: The perihelion motion of Mercury},%
  pdfauthor={C. K\"orber, I. Hammer, J.-L. Wynen, J. Heuer, C. M\"uller and C. Hanhart},%
  pdfkeywords={Numerical Simulations, Teaching, Perihelion Motion of Mercury, General Relativity},%
  pdfsubject={A primer with the goal to teach general concepts of numerical simulations to high school students.  The specific example is the perihelion motion of mercury due to General Relativity.},   % subject of the document
  colorlinks=true,        % true: colored links
  linkcolor=OliveGreen,          % color of internal links (change box color with linkbordercolor)
  citecolor=magenta,        % color of links to bibliography
  filecolor=YellowGreen,      % color of file links
  urlcolor=blue           % color of external links
]{hyperref}
\usepackage[space=true]{accsupp}
% requires the latest version of package accsupp
\newcommand{\copyablespace}{\BeginAccSupp{method=hex,unicode,ActualText=00A0}\ \EndAccSupp{}}

@@

\begin{abstract}
Numerical simulations are playing an increasingly important role in modern science.
In this work it is suggested to use a numerical study of the famous perihelion motion of the planet Mercury (one of the prime observables supporting Einsteins General Relativity) as a test case to teach numerical simulations to high school students.
The paper includes details about the development of the code as well as a discussion of the visualization of the results.
In addition a method is discussed how to estimate the size of the effect a priori. 
\end{abstract}

@@

Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise, e.g., since there are too many particle participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that allows one to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to get in touch with
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \texttt{VPython}.
\end{itemize}
The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their 
discretization is discussed based on Newtons law of gravitation and possible extensions thereof.
Afterwards the visualization of the resulting trajectories using \texttt{VPython} is introduced and applied to the problem at hand. In addition,
 tools are developed to extract the relevant quantity from the result of the simulation.
Finally the principle of dimensional analysis is presented as a tool to cross check, if the results of the simulation are sensible.

We are convinced that in order to excite the students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
While their detailed study needs deeper knowledge in Differential Geometry, for the study outlined here very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
This was already demonstrated in the "Schülderakademie Teilchenphysik", where this course was tested successfully on two groups consisting of in total 24 
German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.

@@

For this procedure to work $\Delta t$ must be sufficiently small.
What this means depends on the process studied.
One way to estimate, if $\Delta t$ is small enough, is to verify, if the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\ . 
\label{eq:check}
\end{equation}
holds, since the next term neglected in Eq.~(\ref{eq:vdef}) reads $(1/2)a(t)(\Delta t)^2$.
Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive the concrete value $\Delta t$ chosen --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For this concrete example the starting point for the force is Newtons law of gravitation
\begin{equation}
\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
In addition $r=|\vec r(t)|$ denotes the distance between sun and Mercury, when we assume that the sun is infinitely more heavy than the planet and located at the center of the coordinate system. 
Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is a pretty good approximation.
For later convenience we introduce the Schwarzschildradius of the sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ ,
\end{equation}
where $c=3\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
That the Schwarzschild radius is also an important quantity to characterize black holes is not relevant for the discussion at hand.
With this Newtons second law reads 
\begin{equation}
	\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, .
\label{eq:newton}
\end{equation}
In general an attractive force that vanishes at large distances leads,  depending on the initial conditions, either to bounded or to open orbits. 
In the latter case the planet simply disappears from the sun, since a given force can capture only bodies with not too large momenta.
The students may study those scenarios within their simulations  by varying the start velocity while keeping the start location fixed.

The bounded orbits that emerge from a potential that scales as $1/r$ are elliptic and fixed in space.
In particular the point of closest approach of the planet to the sun, the perihelion, does not move.
However, as soon as the potential deviates from $1/r$, the perihelion moves.
Because of this, the behavior of the perihelion is a very sensitive probe of the gravitational potential.

The observed perihelion motion of Mercury is nowadays determined as $(574.10\pm 0.65)''$ per
100 earth years~\footnote{The symbol $''$ denotes "arc seconds": 1$''=(1/3600)^o$. }, but it should be noted that the bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of $\delta \Theta_M = (43)''$ in 100 earth years remained unexplained, until Einstein quantified the
predictions of General Relativity to this particular observable~\cite{Einstein}.

To allow for a movement of the perihelion we need to modify the force that appears in Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side with a factor
$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter. 
However, besides $r_S$, which is the characteristic length scale for the gravitational potential
of the sun, one can define one more parameter, which is characteristic for the dynamics of mercury, namely
\begin{equation}
a^2=\frac{\vec L\,^2}{m^2c^2}= \frac{(\vec r\times \dot{\vec r} \, )^2}{c^2} \ ,
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $a^2$ carries dimensions of a length squared.  Therefore we may use  the following ansatz
for the modified potential
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}+\beta\frac{a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand $a^2$ may be estimated from the parameters of mercury at its perihelion ---  $\abs{\vec v_M(0)} = 59$km/s and 
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}, where both $\vec r$ and $\dot{\vec r}$ are perpendicular to each other. 
We thus find that $(a^2/ r_{MS}^2) = 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})=6\cdot 10^{-8}$. 
Therefore one may expect from both correction terms an effect similar in size. Furthermore, any term higher order in either $a$ or $r_S$
should be suppressed by additional six orders of magnitude and thus can be safely neglected.
The parameters $\alpha$ and $\beta$ could now either be extracted from a fit to data or be calculated from the underlying
theory --- here General relativity with the result
\begin{equation}\label{eq:art-prediction}
	\alpha = 0 \, , \quad \beta = 3 \, .
\end{equation}

However, since we have only a single number to study, the angular motion of the perihelion,
we can only fix either $\alpha$ or $\beta$.

%For comparison: the typical distance of the Venus, the neighbor of Mercury, to the sun is $(r_S/r_{VS}) \sim 5 \cdot 10^{-8}$.
%Therefore the expected ratio of residual motion, favoring in the orbit period $T$, would result in
%\begin{equation}
%	\delta \Theta_V \approx \frac{1}{T_V} \delta \Theta_M T_M \frac{r_{MS}}{r_{VS}} \approx 9'' \, ,
%\end{equation}
%where $T_V$ ($T_M$) denotes the length of a venus (mercury) year. This precisely describes the experimental value of $9''$.

@@

In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
particular
the perihelion motion of Mercury.  We select \python{} \cite{Python} as  programming language, because \python{} is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python{} or any other programming language is required, as we explain all the necessary steps to create the simulation.
In appendix \ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
We also provide a working example program that can be used as a template \url{this-upload-url}\todo{@arXiv: online link}.

To start the simulation one needs the "initial" distances and velocities, which are needed as input to the equation of motion [Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})].
As described above, here  one can simplify the problem by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (mercury) in a two-dimensional plane.

Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km and $ \abs{\vec v_M(0)} = 59$km/s\cite{MercuryFactSheet}.
Since the computer does not understand physical units, one has to express each variable in an appropriate unit.
Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".
A useful choice for this problem is given e.g. by expressing distances in $R_0 = 10^{10}$ m and time intervals in days, $T_0 = 1$d, where $d$ refers to earth days, such that one mercury year is given by $T_M=88\,T_0$.
With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration pre-factors become
\begin{eqnarray}
r_{MS}(0)   &=& 4.6 R_0 \, , \quad
v_{M}(0)    = 0.51 \frac{R_0}{T_0} \, ,  \\
a_M(r_{MS}) &=& 1.01 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2} 
\, .
\end{eqnarray}

%
\begin{figure}[htb]
	\centering
	\includegraphics[width=.5\textwidth]{figs/sun_merc.pdf}
	\caption{\label{fig:sun_merc}%
		Sun mercury system with relevant vectors.
		Because mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.%
	}
\end{figure}
%

In \python{}, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.6    # in units of R0
vM0 = 0.51   # in units of R0/T0
c_a = 1.01   # in units of R0/T0**2
TM  = 88.    # in units of T0
rS  = 3.0e-7 # in units of R0
a2  = 8.2e-7 # in units of R0**2
\end{lstlisting}
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python{} module \texttt{VPython}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpythons module visual
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.
However, we first need to fix the time step $\Delta t$ and use the estimate of Eq.~(\ref{eq:check}) as a guidance
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 20
\end{lstlisting}
Here the factor $1/20$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following equations\footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracys.
	Thus, the ordering and exact equations for updating $a_{MS}$, $v_M$ and $r_M$ are not unique, but all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
temp = 1 + alpha * rS / vec_rM_old.mag + beta * a2 / vec_rM_old.mag**2
aMS  = c_a * temp / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \texttt{vector.mag} --- is a property of the vector and can be easily extracted.

It is handy to use \texttt{Python}s \texttt{function}s to embed repeating structures ("DRY" --- Don't Repeat Yourself)
\begin{lstlisting}
# Define the coordinate and velocity update function
def evolve_mercury(vec_rM_old, vec_vM_old, alpha, beta):
	# Compute the strength of the acceleration
	temp = 1 + alpha * rS / vec_rM_old.mag + beta * a2 / vec_rM_old.mag**2
	aMS  = c_a * temp / vec_rM_old.mag**2
	# Multiply by the direction
	vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
	# Update velocity vector
	vec_vM_new = vec_vM_old + vec_aMS * dt
	# Update position vector
	return vec_rM_new, vec_vM_new

# Call the function
vec_rM_new, vec_vM_new = evolve_mercury(vec_rM_old, vec_vM_old, 0.0, 0.0)
\end{lstlisting}
\python{}s syntax enforces a clean programming style: it is necessary that the body of the function is indented (by an arbitrary but consistent amount of \texttt{space}s or \texttt{tab}s) relative to the definition statement of the function.
Furthermore, \python{} is an Interpreter language.
Each line of the code is executed when the Interpreter passes it.
For this reason, all the variables defined before the function are global and thus known to the function.
The variables within the function (e.g. \texttt{aMS}, \texttt{vec\_aMS}, $\dots$) are local and do not exist beyond the scope of the function.

>>>>

\usepackage[dvipsnames]{xcolor}         % Enabling colors by their 'svgnames'
\usepackage[%
  unicode=false,%
  pdftitle={A primer to numerical simulations: The perihelion motion of Mercury},%
  pdfauthor={C. K\"orber, I. Hammer, J.-L. Wynen, J. Heuer, C. M\"uller and C. Hanhart},%
  pdfkeywords={Numerical Simulations, Teaching, Perihelion Motion of Mercury, General Relativity},%
  pdfsubject={A primer with the goal to teach general concepts of numerical simulations to high school students.  The specific example is the perihelion motion of Mercury due to General Relativity.},   % subject of the document
  colorlinks=true,        % true: colored links
  linkcolor=OliveGreen,          % color of internal links (change box color with linkbordercolor)
  citecolor=magenta,        % color of links to bibliography
  filecolor=YellowGreen,      % color of file links
  urlcolor=blue           % color of external links
]{hyperref}
\usepackage[space=true]{accsupp}
% requires the latest version of package accsupp
\newcommand{\copyablespace}{\BeginAccSupp{method=hex,unicode,ActualText=00A0}\ \EndAccSupp{}}

@@

\begin{abstract}
Numerical simulations are playing an increasingly important role in modern science.
In this work it is suggested to use a numerical study of the famous perihelion motion of the planet Mercury (one of the prime observables supporting Einsteins General Relativity) as a test case to teach numerical simulations to high school students.
The paper includes details about the development of the code as well as a discussion of the visualization of the results.
In addition a method is discussed how to estimate the size of the effect a priori that at the same time also can be employed to
double check the results found numerically. 
\end{abstract}

@@

Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise, e.g., since there are too many particle participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that allows one to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to get in touch with
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \texttt{VPython}.
\end{itemize}
The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their 
discretization is discussed based on Newtons law of gravitation and possible extensions thereof.
Afterwards the visualization of the resulting trajectories using \texttt{VPython} is introduced and applied to the problem at hand. In addition,
 tools are developed to extract the relevant quantity from the result of the simulation.
Finally the principle of dimensional analysis is presented as a tool not only to estimate some effect but also to cross check, if the results of the simulation are sensible.

We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
While their detailed study needs deeper knowledge in Differential Geometry, for the study outlined here very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
This was already demonstrated in the "Schülerakademie Teilchenphysik", where this course was tested successfully on two groups consisting of in total 24 
German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017.

@@

For this procedure to work $\Delta t$ must be sufficiently small.
What this means depends on the process studied.
One way to estimate, if $\Delta t$ is small enough, is to verify, if the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\ . 
\label{eq:check}
\end{equation}
holds, since the next term neglected in Eq.~(\ref{eq:vdef}) reads $(1/2)a(t)(\Delta t)^2$.
Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive the concrete value chosen for $\Delta t$ --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For this concrete example the starting point for the force is Newtons law of gravitation
\begin{equation}
\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
In addition $r=|\vec r(t)|$ denotes the distance between sun and Mercury, when we assume that the sun is infinitely more heavy than the planet and located at the center of the coordinate system. 
Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is a pretty good approximation.
For later convenience we introduce the Schwarzschildradius of the sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ ,
\label{rsdef}
\end{equation}
where $c=3\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
That the Schwarzschild radius is also an important quantity to characterize black holes is not relevant for the discussion at hand.
With this Newtons second law reads 
\begin{equation}
	\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, .
\label{eq:newton}
\end{equation}
In general an attractive force that vanishes at large distances leads,  depending on the initial conditions, either to bounded or to open orbits. 
In the latter case the planet simply disappears from the sun, since a given force can capture only bodies with not too large momenta.
The students may study those scenarios within their simulations  by varying the start velocity while keeping the start location fixed.

The bounded orbits that emerge from a potential that scales as $1/r$ are elliptic and fixed in space.
In particular the point of closest approach of the planet to the sun, the perihelion, does not move.
However, when a potential that vanishes for $r\to \infty$ deviates from $1/r$, the perihelion moves.
Because of this, the behavior of the perihelion is a very sensitive probe of the gravitational potential.

The observed perihelion motion of Mercury is nowadays determined as 
$$(574.10\pm 0.65)'' \ \mbox{per
100 earth years}$$
where the symbol $''$ denotes "arc seconds": 1$''=(1/3600)^o$. The bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of 
\begin{equation}\delta \Theta_M = (43)'' \ \mbox{in 100 earth years}\label{delT}\end{equation}
 remained unexplained, until Einstein quantified the
predictions of General Relativity to this particular observable~\cite{Einstein}.

To allow for a movement of the perihelion we need to modify the force that appears in Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side with a factor
$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter. 
However, besides $r_S$, which is the characteristic length scale for the gravitational potential
of the sun, one can define one more parameter, which is characteristic for the dynamics of Mercury, namely
\begin{equation}
a^2=\frac{\vec L\,^2}{m^2c^2}= \frac{(\vec r\times \dot{\vec r} \, )^2}{c^2} \ ,
\label{a2def}
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $a^2$ carries dimensions of a length squared.  Therefore we may use  the following ansatz
for the modified potential
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}+\beta\frac{a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand $a^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is 
 $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59$~km/s (here we
already indicate that below we start the simulation at the perihelion) and the closest distance between sun and Mercury is
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other. 
We thus find that $(a^2/ r_{MS}^2) = 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})=6\cdot 10^{-8}$. 
Therefore (for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}) one may expect from both correction terms an effect similar in size. Furthermore, any term higher order in either $a$ or $r_S$
should be suppressed by additional seven orders of magnitude and thus can be safely neglected.
The parameters $\alpha$ and $\beta$ can be extracted from a fit to data --- this will be done below. 
However, since we have with the perihelion shift (see Eq.~(\ref{delT})) only a single number to study,
we can only fix either $\alpha$ or $\beta$ (or a linear combination thereof).
At the same time the parameters can also be calculated from the assumed underlying theory and the actual values
are in general specific for this theory. It turns out that
General relativity gives~\cite{Einstein}
\begin{equation}\label{eq:art-prediction}
	\alpha = 0 \, , \quad \beta = 3 \, .
\end{equation}
Thus we may also use the simulation explained below to calculate the perihelion motion of Mercury from the input
values given in Eq.~(\ref{eq:art-prediction}).

%For comparison: the typical distance of the Venus, the neighbor of Mercury, to the sun is $(r_S/r_{VS}) \sim 5 \cdot 10^{-8}$.
%Therefore the expected ratio of residual motion, favoring in the orbit period $T$, would result in
%\begin{equation}
%	\delta \Theta_V \approx \frac{1}{T_V} \delta \Theta_M T_M \frac{r_{MS}}{r_{VS}} \approx 9'' \, ,
%\end{equation}
%where $T_V$ ($T_M$) denotes the length of a venus (Mercury) year. This precisely describes the experimental value of $9''$.

@@

In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
particular
the perihelion motion of Mercury.  We select \python{} \cite{Python} as  programming language, because \python{} is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python{} or any other programming language is required, as we explain all the necessary steps to create the simulation.
In  \ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
We also provide a working example program that can be used as a template \url{this-upload-url}\todo{@arXiv: online link}.

To start the simulation one needs the "initial" distance and velocity [Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})].
As described above, here  one can simplify the problem by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (Mercury) in a two-dimensional plane.

Since the trajectories do not depend on where on the orbit of Mercury the simulation is started, we 
use as initial (for $t=0$) parameters the values at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$~km and $ \abs{\vec v_M(0)} = 59$~km/s\cite{MercuryFactSheet}.
Since the computer does not understand physical units, one has to express each variable in an appropriate unit.
Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range", e.g., 
by expressing distances in $R_0 = 10^{10}$ m and time intervals in days, $T_0 = 1$d, where $d$ refers to earth days, such that one Mercury year is given by $T_M=88\,T_0$.
With this choice, the initial distance of Mercury to the sun, the size of the initial velocity of Mercury and the acceleration pre-factors become
\begin{eqnarray}
r_{MS}(0)   &=& 4.6 R_0 \, , \quad
v_{M}(0)    = 0.51 \frac{R_0}{T_0} \, ,  \\
a_M(r_{MS}) &=& 1.01 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2} 
\, .
\end{eqnarray}

%
\begin{figure}[htb]
	\centering
	\includegraphics[width=.5\textwidth]{figs/sun_merc.pdf}
	\caption{\label{fig:sun_merc}%
		Sun Mercury system with relevant vectors.
		Because Mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.%
	}
\end{figure}
%

In \python{}, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.6    # in units of R0
vM0 = 0.51   # in units of R0/T0
c_a = 1.01   # in units of R0/T0**2
TM  = 88.    # in units of T0
rS  = 3.0e-7 # in units of R0
a2  = 8.2e-7 # in units of R0**2 \ ,
\end{lstlisting}
where the last two quantities refer to the Schwarzschildradius of the sun and the parameter $a^2$ defined in Eqs. (\ref{rsdef}) and (\ref{a2def}), respectively.
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python{} module \texttt{VPython}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of Mercury is perpendicular to the vector which connects Mercury and sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpythons module visual
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0) \ .
\end{lstlisting}
To fix the time step $\Delta t$ we use the estimate of Eq.~(\ref{eq:check}) as a guidance to write
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 20 \ .
\end{lstlisting}
Here the factor $1/20$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following equations\footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracies.
	Thus, the ordering and exact equations for updating $a_{MS}$, $v_M$ and $r_{MS}$ are not unique, but all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
temp = 1 + alpha * rS / vec_rM_old.mag + beta * a2 / vec_rM_old.mag**2
aMS  = c_a * temp / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \texttt{vector.mag} --- is a property of the vector and can be easily extracted.

It is handy to use \texttt{Python}s \texttt{function}s to embed repeating structures ("DRY" --- Don't Repeat Yourself)
\begin{lstlisting}
# Define the coordinate and velocity update function
def evolve_Mercury(vec_rM_old, vec_vM_old, alpha, beta):
	# Compute the strength of the acceleration
	temp = 1 + alpha * rS / vec_rM_old.mag + beta * a2 / vec_rM_old.mag**2
	aMS  = c_a * temp / vec_rM_old.mag**2
	# Multiply by the direction
	vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
	# Update velocity vector
	vec_vM_new = vec_vM_old + vec_aMS * dt
	# Update position vector
	return vec_rM_new, vec_vM_new

# Call the function
vec_rM_new, vec_vM_new = evolve_Mercury(vec_rM_old, vec_vM_old, 0.0, 0.0)
\end{lstlisting}
\python{}s syntax enforces a clean programming style: it is necessary that the body of the function is indented (by an arbitrary but consistent amount of \texttt{space}s or \texttt{tab}s) relative to the definition statement of the function.
Furthermore, \python{} is an Interpreter language.
Each line of the code is executed when the Interpreter passes it.
For this reason, all the variables defined before the function are global and thus known to the function.
The variables within the function (e.g. \texttt{aMS}, \texttt{vec\_aMS}, $\dots$) are local and do not exist beyond the scope of the function.

Finally, we can describe the evolution by a \texttt{while}-loop
\begin{lstlisting}%[float,floatplacement=H]
t     = 0.0
alpha = 0.0
beta  = 0.0
# Execute the loop as long as t < 2*TM
while t < 2*TM:
	vec_rM, vec_vM = evolve_Mercury(vec_rM, vec_vM, alpha, beta)
	t = t + dt
\end{lstlisting}
where for the start we set the parameter $\alpha=0 = \beta$ in order first study the properties of the pure 
$1/r^2$ force.
Note the required indent of the loop structure similar to the indent of a function.
In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values, 
which directly overwrites the previous values to enter the next iteration.
The total runtime \texttt{2*TM} is the amount of "virtual" days the simulations should run.
To describe at least one full orbital period, this time needs to be larger than $T_M$.
With the previous choice for \texttt{dt}, this corresponds to roughly $N_T \approx 2 \cdot 10^3$ evolution steps.
Note that the exact time it takes for Mercury to complete a full revolution depends on the initial coordinates and velocities, the accuracy of the computation
(controlled by the value of $\Delta t$) as well as the computing power employed.
One should encourage the students to analyze this in the beginning.
